We are using an Apple MacBookPro (mid-2009 model) machine running Mac OS 10.8.5. The specs of the machine are described below. Some of the following information could be directly found through the system using the \texttt{system\_profiler} command. Some of this information could only be derived from online documentation. In all cases, the origin of the derivation of the characteristics is shown.

\subsection{Processor}
The machine's processor description was obtained from the Apple ~\cite{appleProc} and Intel ~\cite{intelProc} websites. 
The machine possesses a single Intel Wolfdale-3M processor (model E7300) dual core processor. According to Wikipedia ~\cite{WikiWolfdale}, the Wolfdale-3M processor uses the Intel Penryn ~\cite{WikiPenryn} micro-architecture. 

\begin{itemize}
\item{Each core has a clock frequency of 2.66 GHz \footlabel{sysprof}{This information was derived from the system\_profiler.}.}
\item{The L1 cache size is 64kB per core \footlabel{penryn}{This information was derived from the Penryn documentation.}.}
\item{The L2 cache size is 3MB \footref{sysprof}.}
\item{The L3 cache size is 8MB \footref{penryn}}
\item{The processor uses a 64 bit instruction set \footref{sysprof}.}
\end{itemize}

\subsection{Buses}

The machine uses the Intel terminology for buses. Data transfer between the CPU to the memory is done through a 1066 Mhz  front side bus (FSB) \footref{sysprof}. The machine uses four PCI Express 2.0 buses for I/O purposes. Two of them are 16 lanes, two of them are 4 lanes (each lane has a throughput of 250-500 MB/s) ~\cite{AppleBus}.

\subsection{Memory}

The machine possesses two DDR3 2GB memory banks. Both banks use the memory bus mentioned earlier (FSB). The memory banks support but do not use error correcting codes (ECC) \footref{sysprof}. 

\subsection{Storage}

The machine uses a Hitachi HSTS545032B9SA02 drive ~\cite{StarMicro}. It is a rotational hard drive which uses the SATA protocol. It uses a journaled HFS+ file system \footref{sysprof}. Its rotational delay is 11ms and has a seek time of 12 ms (5400RPM). Finally, its controller cache size is 8MB \footlabel{hitachi}{This information was derived from the StarMicro documentation.}. Total disk size is 320GB.

\subsection{I/O Network}
The machine possesses two USB device ports, one firewire port, one Bluetooth port, one wifi port, one 1000 Base T ethernet port. The ethernet bandwidth (network card) is 1GB/s. All of those ports use the aforementioned PCI buses to communicate with the memory \footref{sysprof}.

\subsection{Operating System}
The machine uses the Mac OS X operating system, version 10.8.5 (12F45). It uses the Darwin 12.5.0 kernel and uses secure virtual memory (swap files are encrypted before being written to disk)\footref{sysprof}.